---
title: "ApplicationofMachinelearninginSurv"
author: "Tshepiso Mokoena"
date: "2025-11-19"
output:
  word_document: default
  html_document: default
---
#install.packages(c(
  "survival",
  "survminer",
  "dplyr",
  "caret",
  "glmnet",
  "xgboost",
  "boot",
  "pROC",
  "survcomp",
  "randomForestSRC",
  "survivalmodels",
  "reticulate","riskRegression","timeROC","pec","readr"))
  
#The dataset contains survival information for patients with advanced lung cancer, including both clinical and demographic variables. Each record represents an individual patient, with time indicating survival duration in days and status denoting the event indicator (0 = censored/alive, 1 = dead). Patient demographics include age in years and sex coded as 1 for male and 2 for female. Functional status is captured using two physician-rated measures: ph.ecog, the Eastern Cooperative Oncology Group (ECOG) performance score ranging from 0 (asymptomatic) to 4 (bedbound), and ph.karno, the Karnofsky performance score (0–100, higher values indicating better health). Patients also self-assessed their functional status via pat.karno (0–100). Additional lifestyle and clinical covariates include meal.cal, the average daily caloric intake, and wt.loss, the weight lost in the last six months in pounds. 

```{r Load packages}

#install.packages(c("survival",
"survminer",
"dplyr",
"caret",
"glmnet",
"xgboost",
"boot",
"pROC",
"survcomp",
"randomForestSRC",
"survivalmodels",
"reticulate","riskRegression","timeROC","pec","readr"))
```

```{r}

library(survival)   
library(survminer) 
library(readr)
library(dplyr)        
library(caret)           # Machine learning workflow: data splitting, cross-validation,preprocessing
library(glmnet)          # Penalized regression models: Lasso, Ridge, and Elastic Net for linear,
library(xgboost)         # Gradient boosting framework 
library(boot)            # Bootstrap resampling and confidence interval estimation
library(pROC)            # Receiver Operating Characteristic (ROC) analysis and AUC computation
library(survcomp)        # Survival model evaluation: concordance index, performance metrics
library(randomForestSRC) # Random survival forests for right-censored survival data
library(survivalmodels)  # Deep learning survival models (e.g., DeepSurv) via reticulate/Pycox
library(reticulate)      # Interface to Python from R: required for Pycox/DeepSurv integration
library(riskRegression)  #  Score(), calibration plots, IPCW-adjusted metrics
library(timeROC)         # Time-dependent ROC curves and AUC for censored survival data
library(pec)             # Prediction error curves, calibration, and model validation
```

```{r Data}
set.seed(123)
data_clean <- read.csv("D:/OneDrive/Desktop/ML Short Course/completed_data1.csv")
str(data_clean)
summary(data_clean)

```
```{r}


cont_vars <- c("age","ph.karno","pat.karno","meal.cal","wt.loss")

# Shapiro-Wilk test
for (v in cont_vars) {
  cat("\nVariable:", v, "\n")
  print(shapiro.test(data_clean[[v]]))
}

# Histograms + Density + QQ plots
library(ggplot2)
library(gridExtra)


plot_list <- list()
for (v in cont_vars) {
  df <- data.frame(x = data_clean[[v]])
  
  # Histogram 
  p1 <- ggplot(df, aes(x)) +
    geom_histogram(bins = 20, fill = "skyblue", color = "black") +
    labs(title = paste("Histogram:", v), x = v, y = "Count")
  
  # Q-Q plot
  p2 <- ggplot(df, aes(sample = x)) +
    stat_qq() + stat_qq_line(color = "red") +
    labs(title = paste("Q-Q Plot:", v))
  
  plot_list[[v]] <- gridExtra::arrangeGrob(p1, p2, ncol = 2)
}


gridExtra::grid.arrange(grobs = plot_list, ncol = 1)

data_clean %>%
  group_by(status) %>%
  summarise(
    Median_time = median(time, na.rm = TRUE),
    Q1_time = quantile(time, 0.25, na.rm = TRUE),
    Q3_time = quantile(time, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

data_clean %>%
  group_by(status) %>%
  summarise(
    Median_age = median(age, na.rm = TRUE),
    Q1_age = quantile(age, 0.25, na.rm = TRUE),
    Q3_age = quantile(age, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

```


```{r}
# Outcomes + features
outcomes_analysis <-data_clean %>%
  dplyr::mutate(
    surv_time  = time,
    surv_event = status,
    death_100d   = ifelse(time <= 100  & status == 1, 1, 0),
    death_250d  = ifelse(time <= 250 & status == 1, 1, 0),
    death_400d  = ifelse(time <= 400 & status == 1, 1, 0)
  )

predictors <- c("age","sex","ph.ecog","ph.karno","pat.karno","meal.cal","wt.loss")

```

#The make_design_matrix function converts selected columns from a data frame into a numeric matrix for modeling. It uses model.matrix() to handle factor variables via one-hot encoding and dynamically builds the formula from the predictors vector.
```{r Unified design matrix builder}
# Unified design matrix builder
make_design_matrix <- function(df, predictors) {
  mm <- model.matrix(
    as.formula(paste("~", paste(predictors, collapse = " + "))),
    data = df
  )
  mm[, -1, drop = FALSE]  # drop intercept
}

```
# ============================================================
# Nested CV setup
# ============================================================
```{r Define Fold Counts}


set.seed(123)

n_outer_folds <- 5 #Used to estimate model performance on unseen data
n_inner_folds <- 3#Used for hyperparameter tuning within each outer fold.
```

```{r Create Outer Folds}

#stratified splitting
outer_folds <- createFolds(outcomes_analysis$surv_event, k = n_outer_folds)



nested_cv_results <- data.frame()



```
# ============================================================
# Outer loop - fit models only
# ============================================================

#Training data (train_data) is used for model development — including inner CV, hyperparameter tuning, and fitting.

#Test data (test_data) is used only for final model perf Outcome extraction, scoring, plots

#test_X	Numeric predictors (matrix)	- Predictions from trained models
```{r}

#Iterates(full training + testing cycle.) over each outer fold 1 to 5.

for (fold_i in seq_len(n_outer_folds)) {

  cat(sprintf("\n--- Outer Fold %d/%d ---\n", fold_i, n_outer_folds))
  
  test_idx  <- outer_folds[[fold_i]]
  train_idx <- setdiff(seq_len(nrow(outcomes_analysis)), test_idx)
  
  train_data <- outcomes_analysis[train_idx, ]
  test_data  <- outcomes_analysis[test_idx, ] #held-out test set
  
  train_X <- make_design_matrix(train_data, predictors )
  test_X  <- make_design_matrix(test_data, predictors )
  
  train_surv <- Surv(train_data$surv_time, train_data$surv_event) #Survival outcome used to fit models
} 
  fold_results <- data.frame()
```

```{r Model Development inside the outer fold.}
  
  # ============================================================
  # LASSO Cox
  # ============================================================
  tryCatch({
    lasso_model <- cv.glmnet(train_X, train_surv, family="cox", alpha=1, nfolds=n_inner_folds)
    lasso_lp <- as.numeric(predict(lasso_model, newx=test_X, s="lambda.min", type="link"))
    
    fold_results <- rbind(fold_results,
                          data.frame(Model="Lasso", Fold=fold_i))
  })
  
  # ============================================================
  # Ridge Cox
  # ============================================================
  tryCatch({
    ridge_model <- cv.glmnet(train_X, train_surv, family="cox", alpha=0, nfolds=n_inner_folds)
    ridge_lp <- as.numeric(predict(ridge_model, newx=test_X, s="lambda.min", type="link"))
    
    fold_results <- rbind(fold_results,
                          data.frame(Model="Ridge", Fold=fold_i))
  })
  
  # ============================================================
  # Elastic Net Cox
  # ============================================================
  tryCatch({
    elastic_model <- cv.glmnet(train_X, train_surv, family="cox", alpha=0.5, nfolds=n_inner_folds)
    elastic_lp <- as.numeric(predict(elastic_model, newx=test_X, s="lambda.min", type="link"))
    
    fold_results <- rbind(fold_results,
                          data.frame(Model="ElasticNet", Fold=fold_i))
  })
  
  # ============================================================
  # Random Survival Forest
  # ============================================================
  tryCatch({
    rsf_model <- rfsrc(
      Surv(surv_time, surv_event) ~ .,
      data = train_data[, c("surv_time", "surv_event",predictors)],
      ntree = 100 )
    rsf_pred <- predict(rsf_model, newdata = test_data[, c("surv_time", "surv_event", predictors)])
    rsf_lp   <- rowSums(rsf_pred$chf)
    
    fold_results <- rbind(fold_results,
                          data.frame(Model="RSF", Fold=fold_i))
  })
  
  
  # ============================================================
  # XGBoost Cox
  # ============================================================
  tryCatch({
    
    dtrain <- xgb.DMatrix(data  = as.matrix(train_X),label = train_data$surv_time)
    
    xgb_model <- xgb.train(
      params = list(
        objective = "survival:cox",
        eval_metric = "cox-nloglik",
        eta = 0.05,
        max_depth = 3,
        subsample = 0.8,
        colsample_bytree = 0.8
      ),
      data    = dtrain,
      nrounds = 300,
      verbose = 0
    )
    
    dtest <- xgb.DMatrix(data = as.matrix(test_X))
    xgb_lp <- as.numeric(predict(xgb_model, dtest)) #linear predictors
    
    fold_results <- rbind(fold_results,
                          data.frame(Model="XGBoost", Fold=fold_i))
    
  }, error = function(e) {
    message("XGBoost failed on fold ", fold_i, ": ", e$message)
  })
  

 fold_results
```


```{r  models_fold}

models_fold <- list(
    
    Lasso   = lasso_model,
    Ridge   = ridge_model,
    Elastic = elastic_model,
    RSF     = rsf_model,
    XGB     = xgb_model
    
  )

  fold_results <- models_fold

```
# ============================================================
# Time-dependent evaluation on test folds
# ============================================================
```{r #  Cindex}
library(boot)

#The concordance index (C-index) is a widely used metric in survival analysis that measures the discriminative ability of a model, how well the model can correctly rank individuals according to risk.C-index	Interpretation
#0.5	Model performs no better than random chance
#1.0	Perfect discrimination; all high-risk patients fail before low-risk patients
#<0.5	Predictions are worse than random 

#reusable function to bootstrap the C-index.
bootstrap_cindex_lp <- function(data, lp, R = 500, seed = 123) {
  set.seed(seed)
  
  boot_stat <- function(d, i) {
  concordance.index(lp[i], surv.time = d$surv_time[i], surv.event = d$surv_event[i])$c.index
  }  #Returns C-index for that bootstrap sample.
  
  boot(data = data, statistic = boot_stat, R = R)
}
boot_lasso_cindex   <- bootstrap_cindex_lp(test_data, lasso_lp)
boot_ridge_cindex   <- bootstrap_cindex_lp(test_data, ridge_lp)
boot_elastic_cindex <- bootstrap_cindex_lp(test_data, elastic_lp)
boot_rsf_cindex    <- bootstrap_cindex_lp(test_data, rsf_lp)
boot_xgb_cindex     <- bootstrap_cindex_lp(test_data, xgb_lp)
#boot_deep    <- bootstrap_cindex_lp(deep_test, deep_lp)


cindex_df <- data.frame(
  Model = c("Lasso", "Ridge", "ElasticNet", "RSF", "XGB"),
  Cindex = c(
    boot_lasso_cindex$t0,
    boot_ridge_cindex$t0,
    boot_elastic_cindex$t0,
    boot_rsf_cindex$t0,
    boot_xgb_cindex$t0
  ),
  LowerCI = c(
    boot.ci(boot_lasso_cindex, type="perc")$percent[4],
    boot.ci(boot_ridge_cindex, type="perc")$percent[4],
    boot.ci(boot_elastic_cindex, type="perc")$percent[4],
    boot.ci(boot_rsf_cindex, type="perc")$percent[4],
    boot.ci(boot_xgb_cindex, type="perc")$percent[4]
  ),
  UpperCI = c(
    boot.ci(boot_lasso_cindex, type="perc")$percent[5],
    boot.ci(boot_ridge_cindex, type="perc")$percent[5],
    boot.ci(boot_elastic_cindex, type="perc")$percent[5],
    boot.ci(boot_rsf_cindex, type="perc")$percent[5],
    boot.ci(boot_xgb_cindex, type="perc")$percent[5]
  )
)

print(cindex_df)

```
#The estimated C-index values for the models were 0.586, 0.590, 0.616, 0.609, and 0.561, with standard errors ranging from 0.055 to 0.060. Among these, (Elastic Net Cox) achieved the highest C-index of 0.616, indicating that it has the best predictive discrimination for the outcome. This suggests that the Elastic Net Cox model outperforms Lasso, Ridge, Random Survival Forest, and XGBoost in terms of concordance, making it the preferred model based on C-index.

```{r Penalized brier}

#In survival analysis, it evaluates how well a model predicts the probability of an event  at a specific time.
time_points <- c(100, 250,400) #Evaluation of predictions

feature_names <- colnames(train_X)

  
  
  # --- Wrap models ---
  wrap_model <- function(fit) {
    structure(list(fit = fit, lambda = "lambda.min", feature_names = predictors), class = "glmnet_surv_wrapper")
  }
  
  lasso_wrapped   <- wrap_model(lasso_model)
  ridge_wrapped   <- wrap_model(ridge_model)
  elastic_wrapped <- wrap_model(elastic_model)
  
  # --- Define prediction methods ---
predictSurvProb.glmnet_surv_wrapper <- function(object, newdata, times, ...) {
  x_new <- model.matrix(reformulate(predictors), data = newdata)[, -1]
  

  
  # Reorder to match training
  x_new <- x_new[, object$feature_names, drop = FALSE]
  
  lp <- as.numeric(predict(object$fit, newx = x_new, s = object$lambda, type = "link"))
  
  base_cox <- coxph(Surv(surv_time, surv_event) ~ 1, data = newdata)
  base_haz <- basehaz(base_cox, centered = FALSE)
  H0 <- approx(base_haz$time, base_haz$hazard, xout = times, rule = 2)$y
  
  surv_probs <- outer(exp(lp), H0, function(r, h) exp(-r * h))
  colnames(surv_probs) <- paste0("t_", times)
  surv_probs
}

  
  predictRisk.glmnet_surv_wrapper <- function(object, newdata, times, ...) {
    1 - predictSurvProb(object, newdata = newdata, times = times, ...)
  }
  
  # --- Bootstrap Brier function ---
  bootstrap_brier_glmnet <- function(data, indices, model_wrapped, model_name, times) {
    d <- data[indices, ]
    Score(
      object = setNames(list(model_wrapped), model_name),
      formula = Surv(surv_time, surv_event) ~ 1,
      data = d,
      times = times,
      metrics = "brier",
      cens.model = "cox",
      conservative = TRUE,
      null.model = FALSE
    )$Brier$score$Brier
  }
  
  # --- Run bootstrap ---
  set.seed(123)
  boot_lasso <- boot(test_data, function(data, i) {
    bootstrap_brier_glmnet(data, i, lasso_wrapped, "Lasso", time_points)
  }, R = 200)
  
  boot_ridge <- boot(test_data, function(data, i) {
    bootstrap_brier_glmnet(data, i, ridge_wrapped, "Ridge", time_points)
  }, R = 200)
  
  boot_elastic <- boot(test_data, function(data, i) {
    bootstrap_brier_glmnet(data, i, elastic_wrapped, "ElasticNet", time_points)
  }, R = 200)
  
  # --- Extract CI ---
  extract_ci <- function(boot_obj) {
    lapply(seq_along(time_points), function(j) {
      ci <- boot.ci(boot_obj, type = "perc", index = j)
      c(lower = ci$percent[4], upper = ci$percent[5])
    })
  }
  


 summarize <- function(boot_obj, ci_list, model_name) {
  n <- length(boot_obj$t0)
  stopifnot(n == length(time_points))  # fail early if mismatch
  
  data.frame(
    Fold    = rep(fold_i, n),
    Model   = rep(model_name, n),
    Time    = time_points,
    Brier   = as.numeric(boot_obj$t0),
    LowerCI = as.numeric(sapply(ci_list, `[[`, "lower")),
    UpperCI = as.numeric(sapply(ci_list, `[[`, "upper")),
    stringsAsFactors = FALSE
  )
}
 

results_list <- list(
  summarize(boot_lasso, extract_ci(boot_lasso), "Lasso"),
  summarize(boot_ridge, extract_ci(boot_ridge), "Ridge"),
  summarize(boot_elastic, extract_ci(boot_elastic), "ElasticNet")
)


```


#Elastic Net Cox achieved Brier scores of 0.166, 0.237, and 0.204 at 100, 250, and 400 days, respectively, with confidence intervals

#Lasso Cox  similar scores, slightly lower at 100 and 250 days (0.164 and 0.233) and comparable at 400 days (0.204), indicating similar predictive performance to Elastic Net.

#Ridge Cox showed higher Brier scores at all time points (0.275, 0.345, and 0.262), suggesting lower prediction accuracy compared to Elastic Net and Lasso.


```{r time dep brier  RSF}

bootstrap_brier_riskreg <- function(data, indices, rsf_model, times) {
  d <- data[indices, ]
  
  # Predict survival probabilities
  surv_probs <- predictSurvProb(rsf_model, newdata = d, times = times)
  
  # Score
  sc <- tryCatch({
    Score(
      object = list(RSF = surv_probs),
      formula = Surv(surv_time, surv_event) ~ 1,
      data = d,
      times = times,
      metrics = "brier",
      cens.model = "cox",
      conservative = TRUE,
      null.model = FALSE
    )
  }, error = function(e) NULL)
  
  if (is.null(sc)) return(rep(NA_real_, length(times)))
  
  sc$Brier$score$Brier[sc$Brier$score$model == "RSF"]
}

set.seed(123)

boot_rsf <- boot(
  data = test_data,
  statistic = function(data, i) {
    bootstrap_brier_riskreg(data, i, rsf_model, time_points)
  },
  R = 200
)

rsf_brier_bootCI <- lapply(seq_along(time_points), function(j) {
  boot.ci(boot_rsf, type = "perc", index = j)
})

rsf_brier <- data.frame( Model = "RSF",
  Time = time_points,
  Brier = boot_rsf$t0,
  LowerCI = sapply(rsf_brier_bootCI, function(x) if(!is.null(x)) x$percent[4] else NA),
  UpperCI = sapply(rsf_brier_bootCI, function(x) if(!is.null(x)) x$percent[5] else NA)
)

rsf_brier
```

```{r  time dep brier xgb}

set.seed(123)

# design matrices with consistent features
formula <- ~ age + sex + ph.ecog + ph.karno + pat.karno + wt.loss
train_X <- model.matrix(formula, data = train_data)[, -1]
test_X  <- model.matrix(formula, data = test_data)[, -1]
feature_names <- colnames(train_X)
colnames(test_X) <- feature_names

# XGBoost Cox model
dtrain <- xgb.DMatrix(data = train_X, label = train_data$surv_time)
setinfo(dtrain, "label_lower_bound", ifelse(train_data$surv_event == 1, train_data$surv_time, NA))
setinfo(dtrain, "label_upper_bound", train_data$surv_time)

xgb_model <- xgb.train(
  params = list(
    objective = "survival:cox",
    eval_metric = "cox-nloglik",
    eta = 0.05,
    max_depth = 3,
    subsample = 0.8,
    colsample_bytree = 0.8
  ),
  data = dtrain,
  nrounds = 300,
  verbose = 0
)

# baseline cumulative hazard from Cox model
cox_base <- coxph(Surv(surv_time, surv_event) ~ 1, data = train_data)
base_haz <- basehaz(cox_base, centered = FALSE)

#  Bootstrap estimation IPCW Brier score
bootstrap_brier_xgb <- function(data, indices, xgb_model, base_haz, times, feature_names) {
  d <- data[indices, ]
  
  x <- model.matrix(~ age + sex + ph.ecog + ph.karno + pat.karno + wt.loss, data = d)[,-1]
  colnames(x) <- feature_names
  dtest <- xgb.DMatrix(data = x)
  
  xgb_lp <- as.numeric(predict(xgb_model, newdata = test_X ))
  H0 <- approx(base_haz$time, base_haz$hazard, xout = times, rule = 2)$y
  
  surv_probs <- outer(exp(xgb_lp), H0, function(r, h) exp(-r * h))
  colnames(surv_probs) <- paste0("t_", times)
  
  sc <- Score(
    object = list(XGB = surv_probs),
    formula = Surv(surv_time, surv_event) ~ 1,
    data = d,
    times = times,
    metrics = "brier",
    cens.model = "cox",
    conservative = TRUE,
    null.model = FALSE
  )
  
  sc$Brier$score$Brier[sc$Brier$score$model == "XGB"]
}

#  Run bootstrap -
set.seed(123)
boot_xgb <- boot(
  data = test_data,
  statistic = function(data, i) {
    bootstrap_brier_xgb(data, i, xgb_model, base_haz, time_points, feature_names)
  },
  R = 200
)

#Extract CI
xgb_ci <- lapply(seq_along(time_points), function(j) {
  ci <- boot.ci(boot_xgb, type = "perc", index = j)
  c(lower = ci$percent[4], upper = ci$percent[5])
})


xgb_summary <- data.frame( Model = "XGB",
  Time    = time_points,
  Brier   = boot_xgb$t0,
  LowerCI = sapply(xgb_ci, `[[`, "lower"),
  UpperCI = sapply(xgb_ci, `[[`, "upper")
)

print(xgb_summary)

```

#ElasticNet demonstrated the best performance in terms of Brier scores, consistently achieving low values across all time points (0.166 at 100 days, 0.237 at 250 days, 0.204 at 400 days), indicating the most accurate probability predictions.
```{r  time dep   AUC all}
library(timeROC)
set.seed(123)
# Bootstrap wrapper for timeROC
boot_auc <- function(data, indices, marker, time_points) {
  d <- data[indices, ]
  roc <- timeROC(
    T = d$surv_time,
    delta = d$surv_event,
    marker = marker[indices],
    cause = 1,
    times = time_points,
    iid = FALSE
  )
  roc$AUC
}

# bootstrap 
bootstrap_auc <- function(test_data, marker, time_points, R = 200, model_name = "Model") {
  set.seed(123)
  
  boot_res <- boot(
    data = test_data,
    statistic = function(d, i) boot_auc(d, i, marker, time_points),
    R = R
  )
  
  roc_full <- timeROC(
    T = test_data$surv_time,
    delta = test_data$surv_event,
    marker = marker,
    cause = 1,
    times = time_points,
    iid = FALSE
  )
   
  ci_list <- lapply(seq_along(time_points), function(j) {
    ci <- tryCatch(boot.ci(boot_res, type = "perc", index = j), error = function(e) NULL)
    if (is.null(ci)) {
      c(lower = NA, upper = NA)
    } else {
      c(lower = ci$percent[4], upper = ci$percent[5])
    }
  })
  
  data.frame(
    Model   = model_name,
    Time    = time_points,
    AUC     = round(roc_full$AUC, 3),
    LowerCI = round(sapply(ci_list, `[[`, "lower"), 3),
    UpperCI = round(sapply(ci_list, `[[`, "upper"), 3)
  )
}

```

```{r  AUC}
set.seed(123)
#AUC measures how well the model distinguishes between patients who experience the event versus those who do not.

#Values closer to 1 indicate better discrimination; 0.5 indicates no better than chance.
#Model linear predictors
model_preds <- list(
  Lasso     = lasso_lp,
  Ridge     = ridge_lp,
  ElasticNet= elastic_lp,
  RSF       = rsf_lp,
  XGB       = xgb_lp

)

# Run bootstrap AUC for all models
auc_results <- lapply(names(model_preds), function(name) {
  bootstrap_auc(test_data, model_preds[[name]], time_points, R = 200, model_name = name)
})

# Combine into one summary table
auc_summary_all <- do.call(rbind, auc_results)

print(auc_summary_all)

```
#ElasticNet performs best overall, with the highest AUC at 250 days (0.711) and strong discrimination at 100 (0.637) and 400 days (0.649). RSF is close behind at 250 days, while Lasso, Ridge, and XGBoost show moderate to lower discrimination.


# Quantile-Based time Selection

#  Extract event times only
#event_times <- test_data$surv_time[test_data$surv_event == 1]

# Choose quantiles ( 25%, 50%, 75%)
#quantile_times <- quantile(event_times, probs = c(0.25, 0.5, 0.75))

#  Round to nearest int
#safe_time_points <- round(as.numeric(quantile_times))

# Times
#safe_time_points

```{r}

library(pROC)

# ROC obj at  100 days
roc_lasso_100 <- roc(test_data$death_100d, lasso_lp, smooth = TRUE, quiet = TRUE)
roc_ridge_100 <- roc(test_data$death_100d, ridge_lp, smooth = TRUE, quiet = TRUE)
roc_enet_100  <- roc(test_data$death_100d, elastic_lp, smooth = TRUE, quiet = TRUE)
roc_rsf_100   <- roc(test_data$death_100d, rsf_lp, smooth = TRUE, quiet = TRUE)
roc_xgb_100   <- roc(test_data$death_100d, xgb_lp, smooth = TRUE, quiet = TRUE)
#roc_deep_100 <- roc(test_data_deep$death_100d, deep_lp, smooth = TRUE, quiet = TRUE)

# ROC obj at 250 days
roc_lasso_250 <- roc(test_data$death_250d, lasso_lp, smooth = TRUE, quiet = TRUE)
roc_ridge_250 <- roc(test_data$death_250d, ridge_lp, smooth = TRUE, quiet = TRUE)
roc_enet_250  <- roc(test_data$death_250d, elastic_lp, smooth = TRUE, quiet = TRUE)
roc_rsf_250   <- roc(test_data$death_250d, rsf_lp, smooth = TRUE, quiet = TRUE)
roc_xgb_250   <- roc(test_data$death_250d, xgb_lp, smooth = TRUE, quiet = TRUE)

#ROC curves at 400 days
roc_lasso_400 <- roc(test_data$death_400d, lasso_lp, smooth = TRUE, quiet = TRUE)
roc_ridge_400 <- roc(test_data$death_400d, ridge_lp, smooth = TRUE, quiet = TRUE)
roc_enet_400  <- roc(test_data$death_400d, elastic_lp, smooth = TRUE, quiet = TRUE)
roc_rsf_400   <- roc(test_data$death_400d, rsf_lp, smooth = TRUE, quiet = TRUE)
roc_xgb_400   <- roc(test_data$death_400d, xgb_lp, smooth = TRUE, quiet = TRUE)


# ROC at 100 days

plot(roc_ridge_100, col = "red", lwd = 2, main = "ROC at 100 Days", xlim = c(1,0), ylim = c(0,1))
plot(roc_lasso_100, col = "pink", lwd = 2, add = TRUE)
plot(roc_enet_100, col = "blue", lwd = 2, add = TRUE)
plot(roc_rsf_100, col = "green", lwd = 2, add = TRUE)
plot(roc_xgb_100, col = "purple", lwd = 2, add = TRUE)

legend("bottomright", legend = c(
  paste0("Lasso (AUC = ", round(auc(roc_lasso_100), 3), ")"),
  paste0("Ridge (AUC = ", round(auc(roc_ridge_100), 3), ")"),
  paste0("ElasticNet (AUC = ", round(auc(roc_enet_100), 3), ")"),
  paste0("RSF (AUC = ", round(auc(roc_rsf_100), 3), ")"),
  paste0("XGB (AUC = ", round(auc(roc_xgb_100), 3), ")")
), col = c("pink", "red", "blue", "green", "purple"), lwd = 2, cex = 0.8, bty = "n", y.intersp = 1.2)

# ROC at 250 days

plot(roc_ridge_250, col = "red", lwd = 2, main = "ROC at 250 Days", xlim = c(1,0), ylim = c(0,1))
plot(roc_lasso_250, col = "pink", lwd = 2, add = TRUE)
plot(roc_enet_250, col = "blue", lwd = 2, add = TRUE)
plot(roc_rsf_250, col = "green", lwd = 2, add = TRUE)
plot(roc_xgb_250, col = "purple", lwd = 2, add = TRUE)

legend("bottomright", legend = c(
  paste0("Lasso (AUC = ", round(auc(roc_lasso_250), 3), ")"),
  paste0("Ridge (AUC = ", round(auc(roc_ridge_250), 3), ")"),
  paste0("ElasticNet (AUC = ", round(auc(roc_enet_250), 3), ")"),
  paste0("RSF (AUC = ", round(auc(roc_rsf_250), 3), ")"),
  paste0("XGB (AUC = ", round(auc(roc_xgb_250), 3), ")")
), col = c("pink", "red", "blue", "green", "purple"), lwd = 2, cex = 0.8, bty = "n")

# ROC at 400 days

plot(roc_ridge_400, col = "red", lwd = 2, main = "ROC at 400 Days", xlim = c(1,0), ylim = c(0,1))
plot(roc_lasso_400, col = "pink", lwd = 2, add = TRUE)
plot(roc_enet_400, col = "blue", lwd = 2, add = TRUE)
plot(roc_rsf_400, col = "green", lwd = 2, add = TRUE)
plot(roc_xgb_400, col = "purple", lwd = 2, add = TRUE)

legend("bottomright", legend = c(
  paste0("Lasso (AUC = ", round(auc(roc_lasso_400), 3), ")"),
  paste0("Ridge (AUC = ", round(auc(roc_ridge_400), 3), ")"),
  paste0("ElasticNet (AUC = ", round(auc(roc_enet_400), 3), ")"),
  paste0("RSF (AUC = ", round(auc(roc_rsf_400), 3), ")"),
  paste0("XGB (AUC = ", round(auc(roc_xgb_400), 3), ")")
), col = c("pink", "red", "blue", "green", "purple"), lwd = 2, cex = 0.8, bty = "n", y.intersp = 1.2)





```
```{r Calibration Plots}
event_list <- list(
  "100" = test_data$death_100d,
  "250" = test_data$death_250d,
  "400" = test_data$death_400d
)


get_calibration_summary <- function(pred_prob, event_t, model_name, t_horizon, plot = FALSE, B = 200) {
  if(length(pred_prob) != length(event_t)) {
    stop("Length mismatch: pred_prob and event_t must be the same.")
  }

  pred_prob <- pmin(pmax(pred_prob, 1e-6), 1 - 1e-6)
  logit_pred <- qlogis(pred_prob)
  df <- data.frame(event = event_t, logit_pred = logit_pred)

  # Original models
  cal_int_model <- glm(event ~ offset(logit_pred), family = binomial, data = df)
  intercept <- coef(cal_int_model)[1]

  cal_slope_model <- glm(event ~ logit_pred, family = binomial, data = df)
  slope <- coef(cal_slope_model)[2]

  # Bootstrap
  boot_intercepts <- numeric(B)
  boot_slopes <- numeric(B)

  set.seed(123)
  for (b in 1:B) {
    idx <- sample(seq_len(nrow(df)), replace = TRUE)
    boot_df <- df[idx, ]

    boot_int <- tryCatch({
      fit <- glm(event ~ offset(logit_pred), family = binomial, data = boot_df)
      if (!fit$converged) NA else coef(fit)[1]
    }, error = function(e) NA)

    boot_slope <- tryCatch({
      fit <- glm(event ~ logit_pred, family = binomial, data = boot_df)
      if (!fit$converged) NA else coef(fit)[2]
    }, error = function(e) NA)

    boot_intercepts[b] <- boot_int
    boot_slopes[b] <- boot_slope
  }

  # Remove NAs and compute percentile CIs
  boot_intercepts <- na.omit(boot_intercepts)
  boot_slopes <- na.omit(boot_slopes)

  int_ci <- quantile(boot_intercepts, probs = c(0.025, 0.975))
  slope_ci <- quantile(boot_slopes, probs = c(0.025, 0.975))

  # Safe binning
  qtiles <- quantile(pred_prob, probs = seq(0, 1, 0.1), na.rm = TRUE)
  if (length(unique(qtiles)) < length(qtiles)) {
    qtiles <- seq(min(pred_prob), max(pred_prob), length.out = 11)
  }
  deciles <- cut(pred_prob, breaks = qtiles, include.lowest = TRUE, labels = FALSE)

  calib_df <- data.frame(
    decile = deciles,
    predicted = pred_prob,
    observed = event_t
  ) %>%
    group_by(decile) %>%
    summarise(
      mean_predicted = mean(predicted),
      mean_observed = mean(observed),
      n = n(),
      .groups = "drop"
    ) %>%
    mutate(Model = model_name, Time_Days = t_horizon)

  if (plot) {
    print(
      ggplot(calib_df, aes(x = mean_predicted, y = mean_observed)) +
        geom_point(size = 2, color = "darkgreen") +
        geom_smooth(method = "loess", se = FALSE, color = "darkgreen") +
        geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
        annotate("text", x = 0.05, y = 0.9,
                 label = sprintf("Slope: %.2f (%.2f–%.2f)\nIntercept: %.2f (%.2f–%.2f)",
                                 slope, slope_ci[1], slope_ci[2],
                                 intercept, int_ci[1], int_ci[2]),
                 hjust = 0, size = 4) +
        labs(title = paste("Calibration:", model_name, "at", t_horizon, "days"),
             x = "Predicted Probability", y = "Observed Proportion") +
        theme_minimal()
    )
  }

  return(list(
    intercept = intercept,
    intercept_CI = round(int_ci, 3),
    intercept_CI_Width = round(diff(int_ci), 3),
    slope = slope,
    slope_CI = round(slope_ci, 3),
    slope_CI_Width = round(diff(slope_ci), 3),
    calib_curve = calib_df
  ))
}


```
```{r}

run_model_calibration <- function(lp, event_list, model_name, B = 200) {
  calib_curves <- list()
  metrics_list <- list()

  for (t in names(event_list)) {
    event_t <- event_list[[t]]
    lp_t <- (lp - min(lp)) / (max(lp) - min(lp))

    result <- get_calibration_summary(
      pred_prob = lp_t,
      event_t = event_t,
      model_name = model_name,
      t_horizon = as.numeric(t),
      plot = FALSE,
      B = B
    )

    calib_curves[[t]] <- result$calib_curve
    metrics_list[[t]] <- list(
      Model = model_name,
      Time_Days = as.numeric(t),
      Intercept = result$intercept,
      Intercept_Lower = result$intercept_CI[1],
      Intercept_Upper = result$intercept_CI[2],
      Intercept_CI_Width = result$intercept_CI_Width,
      Slope = result$slope,
      Slope_Lower = result$slope_CI[1],
      Slope_Upper = result$slope_CI[2],
      Slope_CI_Width = result$slope_CI_Width
    )
  }

  calib_df <- bind_rows(calib_curves)
  metrics_df <- bind_rows(metrics_list) %>%
    distinct(Model, Time_Days, .keep_all = TRUE)

  return(list(
    calib_df = calib_df,
    metrics_df = metrics_df
  ))
}

```


```{r}
plot_calibration_loess <- function(calib_df, metrics_df, model_name, color_map) {
  ggplot(calib_df, aes(x = mean_predicted, y = mean_observed)) +
    geom_point(size = 2, color = color_map[model_name]) +
    geom_smooth(method = "loess", se = FALSE, color = color_map[model_name]) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
    facet_wrap(~Time_Days, nrow = 1) +
    geom_text(
      data = metrics_df,
      aes(x = 0.05, y = 0.9,
          label = sprintf("Slope: %.2f (%.2f–%.2f)\nIntercept: %.2f (%.2f–%.2f)",
                          Slope, Slope_Lower, Slope_Upper,
                          Intercept, Intercept_Lower, Intercept_Upper)),
      inherit.aes = FALSE,
      hjust = 0, size = 4
    ) +
    labs(title = paste(model_name, ": Calibration Plot"),
         x = "Predicted Probability",
         y = "Observed Proportion") +
    theme_minimal()
}

```

```{r}
colors <- c(
 
  "RSF"        = "#7570b3", 
  "GBM"        = "#e7298a",
  "Lasso"      = "#66a61e",
  "Ridge"      = "#e6ab02",
  "ElasticNet" = "#a6761d"
 
)


lasso_results <- run_model_calibration(lasso_lp, event_list, "Lasso")
lasso_plot <- plot_calibration_loess(lasso_results$calib_df, lasso_results$metrics_df, "Lasso", colors)

ridge_results <- run_model_calibration(ridge_lp, event_list, "Ridge")
ridge_plot <- plot_calibration_loess(ridge_results$calib_df, ridge_results$metrics_df, "Ridge", colors)

enet_results <- run_model_calibration(elastic_lp, event_list, "ElasticNet")
enet_plot <- plot_calibration_loess(enet_results$calib_df, enet_results$metrics_df, "ElasticNet", colors)

rsf_results <- run_model_calibration(rsf_lp , event_list, "RSF")
rsf_plot <- plot_calibration_loess(rsf_results$calib_df, rsf_results$metrics_df, "RSF", colors)

gbm_results <- run_model_calibration(xgb_lp, event_list, "GBM")
gbm_plot <- plot_calibration_loess(gbm_results$calib_df, gbm_results$metrics_df, "GBM", colors)



slope_intercept <- bind_rows(
  lasso_results$metrics_df,
  ridge_results$metrics_df,
  enet_results$metrics_df,
  rsf_results$metrics_df,
  gbm_results$metrics_df,

)
slope_intercept



lasso_plot 
ridge_plot 
enet_plot 
gbm_plot 
rsf_plot
```

#Summary: ElasticNet and Lasso showed the best calibration across time points, with intercepts near zero and slopes closer to 1, whereas RSF and GBM tended to over- or under-predict risk and produced slopes below 1, indicating exaggerated risk differences.

#Interpretation of intercepts and slopes:

#Intercept: Measures systematic bias in predictions. An intercept of 0 means the model is neither consistently over- nor under-predicting risk. Positive values indicate underprediction (predicted risks too low), and negative values indicate overprediction (predicted risks too high).

#Slope: Measures how well the predicted risks are spread relative to observed outcomes. A slope of 1 indicates perfect calibration (predicted risks match observed risk differences). Slopes <1 suggest predictions are too extreme (high-risk patients’ risks are overestimated, low-risk patients’ risks are underestimated), while slopes >1 indicate predictions are too moderate.
```{r}
# Extract sparse coefficient vector
elastic_model

optimal_lambda <- elastic_model$lambda.min

coef_obj <- coef(elastic_model, s = optimal_lambda)
coef_obj
```
#sex, ECOG performance status, and patient Karnofsky score—while shrinking age, physician Karnofsky score, meal calories, and weight loss to zero, indicating low predictive contribution.
```{r}
library(ggplot2)


coef_df <- data.frame(
  variable = rownames(coef_obj),
  coefficient = as.numeric(coef_obj)
)

# Remove zero-coefficients (variables not selected)
coef_df <- coef_df[coef_df$coefficient != 0, ]


ggplot(coef_df, aes(x = reorder(variable, coefficient), y = coefficient)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Elastic Net Feature Importance (Coefficients)",
    x = "Variable",
    y = "Coefficient"
  ) +
  theme_minimal(base_size = 14)

```
